---
title: "Explore Wind Data"
date: "`r Sys.Date()`"
author: "Danielle Dempsey"
format: 
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
echo: false
editor_options: 
  chunk_output_type: console
---


```{r}
#| include: false

library(dplyr)
library(ggplot2)
library(plotly)
library(RColorBrewer)

theme_set(theme_light())

v_calculate_components <- function(theta_degrees, v_length = 1) {
  
  v_x <- v_length * sin(theta_degrees * pi / 180)
  v_y <- v_length * cos(theta_degrees * pi / 180)
  
  data.frame(v_x = v_x, v_y = v_y)
}


# v_col <- "black"
# pal <- c("#E7298A", "#1B9E77")

# p <- ggplot() +
#   geom_hline(yintercept = 0, col = "grey50") +
#   geom_vline(xintercept = 0, col = "grey50") +
#   scale_x_continuous("x", limits = c(-1, 1)) +
#   scale_y_continuous("y", limits = c(-1,1)) +
#   coord_fixed(ratio = 1) #+  
#   theme(
#     legend.position = "inside",
#     legend.position.inside =  c(0.15, 0.8)
#   )

v_plot_components <- function(
    theta_degrees, v_length = 1, add_components = TRUE
) {
  
  v <- v_calculate_components(theta_degrees, v_length) %>% 
    mutate(theta_degrees = ordered(theta_degrees, levels = theta_degrees))
  
  if(nrow(v) < 3) {
    n_theta <- 3
  } else n_theta <- nrow(v)
  
  
  p_v <- ggplot() +
    geom_hline(yintercept = 0, col = "grey50") +
    geom_vline(xintercept = 0, col = "grey50") +
    scale_x_continuous("x", limits = c(-1, 1)) +
    scale_y_continuous("y", limits = c(-1,1)) +
    coord_fixed(ratio = 1)+ 
    geom_segment(
      aes(x = 0, y = 0, xend = v[,1], yend = v[,2], col = v[,3]),
      arrow = arrow(length = unit(0.3, "cm"))
    ) +
    scale_colour_manual("theta (degrees)", values = brewer.pal(n_theta, "Dark2"))
  
  if(isTRUE(add_components)) {
    
    p_v <- p_v +
      # x component
      geom_segment(
        aes(x = 0, xend = v[,1], y = v[,2], yend = v[,2]),
        arrow = arrow(length = unit(0.3, "cm")), linetype = 2,
      ) +
      # y component
      geom_segment(
        aes(x = 0, xend = 0, y = 0, yend = v[,2]),
        arrow = arrow(length = unit(0.3, "cm")), linetype = 2
      )
  }
  
  p_v
}

# v <- data.frame(
#   x_start = c(0, 0, 0, 0),
#   y_start = c(0, 0, 0, 0),
#   angle_deg = c(0, 360, 45, 20),
#   length = c(1, 0.5, 1, 0.75)
# ) %>% 
#   mutate(
#     angle_rad = round(angle_deg * pi / 180, digits = 5),
#     x_end = x_start + length * sin(angle_rad),
#     y_end = y_start + length * cos(angle_rad),
#     angle_deg = ordered(angle_deg, levels = angle_deg)
#   # ) 

# v %>% 
#   plot_ly(x = ~x_start,
#           y = ~y_start) %>% 
#   add_markers() %>% 
#   add_annotations(text = ~angle_deg,
#                   showarrow = TRUE,
#                   xref = "x_start", axref = "x_start",
#                   yref = "y_start", ayref = "y_start",
#                   x = ~x_end,
#                   ax = ~x_start,
#                   y = ~y_end,
#                   ay = ~y_start)



```

# Motivation

Consider two vectors, one pointing to 0 degrees and one pointing to 360 degrees.

These vectors both point North. The average should therefore also point North.

```{r}
#| fig-height: 4
#| echo: false

p2
```

Taking a simple average:

```{r}
#| echo: true

v_north <- c(0, 360)

mean(v_north)
```

180 degrees points South!

We need another way to calculate the average direction.

# Trigonometry

## Example 1

Now let's consider a vector of length 1 pointing to the North-East, 45 degrees from vertical.

```{r}
#| fig-height: 3
#| echo: false


v_plot_components


v_45 <- v %>% filter(angle_deg == 45)

p + 
  geom_segment(
    data = filter(v, angle_deg %in% c(45)),
    aes(
      x = x_start, y = y_start, xend = x_end, yend = y_end
    ),
    col = v_col,
    arrow = arrow(length = unit(0.3, "cm"))
  ) 
```

This can be broken down into 2 components: one along the x-axis, and one along the y-axis.

```{r}

plot_vector_components(p, v_45) +
  annotate("text", x = 0.07, y = 0.15, label = "45\u00B0")

```

This makes a right-angle triangle, with our original vector as the hypotenuse and the angle theta between the y-axis and our vector.

Now we get to use our high school trigonometry: SOH CAH TOA!

### x component

We use the sine to calculate the length of the x component, since it is opposite from the angle theta (SOH).

$v_x = v_{length} * sin(theta)$

This equation assumes theta is in radians.

Recall: $rad = degrees * pi / 180$

```{r}

v_x = 1 * sin(45 * pi / 180)

v_x
```

### y component

We use the cosine to calculate the length of the y component, since it is adjacent to the angle theta (CAH).

$v_y = v_{length} * cos(theta)$


```{r}

v_y = 1 * cos(45 * pi / 180)

v_y
```

Note that `v_x` and `v_y` have the same length. This is because theta = 45 degrees.

If you remember your unit circle, you might have also noticed the component lengths correspond to sqrt(2) / 2.

## Example 2

Now let's consider a vector of length 0.75 pointing 20 degrees from vertical.


```{r}

plot_vector_components(p, filter(v, angle_deg == 20)) +
  annotate("text", x = 0.07, y = 0.35, label = "20\u00B0")

```


```{r}
v_x = 0.75 * sin(20 * pi / 180)

v_y = 0.75 * cos(20 * pi / 180)

v_x
v_y
```
