---
format: 
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
params:
  round_int: round_int
echo: false
dpi: 600
editor_options: 
  chunk_output_type: console
---

```{r}
#| include: false

library(data.table)
library(dplyr)
library(DT)
library(ggplot2)
library(here)
library(htmltools)
library(leaflet)
library(lubridate)
library(plotly)
library(RColorBrewer)
library(stringr)
library(tidyr)

source(here("functions/subchunkify.R"))
source(here("functions/helpers.R"))

theme_set(theme_light())

round_int <- "6 hours"
#round_int <- params$round_int

# from west to east
stations <- c("YARMOUTH RCS", "LUNENBURG", "BEAVER ISLAND (AUT)", "HART ISLAND (AUT)" )

dat <- fread(
  here(paste0("data/climate_data_", gsub(" ", "_", round_int),  ".csv")),  
  data.table = FALSE
) %>% 
  filter(variable == "bui_m3_s") %>% 
  mutate(
    station = ordered(station, levels = stations),
    value = round(value, digits = 0)
    )

met_locations <- fread(
  here("data/climate_station_locations.csv"), data.table = FALSE
) 
 
get_pal <- colorRampPalette(brewer.pal(8, "Set2"))
pal <- get_pal(length(unique(dat$variable)))

```

# Explore Bakun's Upwelling Index

`r Sys.Date()`

The Bakun Upwelling Index (BUI) is an upwelling index based on Ekman transport, typically expressed in units of $m^3/s$ per 100 m of coastline. Positive values indicate upwelling and negative values indicate downwelling.

$BUI = M * cos(\phi)$

$M = \tau / (\rho_{water} * f)$

$\tau = \rho_{air} * C_D * U^2$

  - $\phi$ is the difference between the coastline angle and the direction the wind is blowing towards (both measured clockwise from the North)
  - $U$ is the wind speed in m/s
  - $\rho_{air}$: air density (1.28 kg/m$^3$) 
  - $\rho_{water}$: sea water density (1026 kg/m$^3$)
  - $C_D$: the wind drag coefficient (0.0015)
  - $f$: Coriolis parameter, which is dependent on latitude[^1]
  
The coastline angle was 60° for BEAVER ISLAND (AUT), HART ISLAND (AUT), and LUNENBURG (following Socrasi et al, 2025). A coastline angle of 330° was used for YARMOUTH RCS.

[^1]: $f = 2\Omega\sin\phi$, where $\Omega$: Earth's rotation rate (7.292x10$^{-5}s^{-1}$) and $\phi$ is the latitude.

## Station Locations

```{r}
#| fig-height: 5

leaflet() %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addCircleMarkers(
    data = met_locations,
    lng = ~longitude,
    lat = ~latitude, label = ~station,
    weight = 1, fillOpacity = 0.75, radius = 5
  ) %>%
  addScaleBar(
    position = "bottomleft",
    options = scaleBarOptions(imperial = FALSE)
  )
```

## Time Series


```{r}
#| results: asis

summer <- data.frame(
  x_start = as_datetime(
    c("2019-07-01",
      "2020-07-01", "2021-07-01", "2022-07-01", "2023-07-01", "2024-07-01", "2025-07-01")),
  x_end = as_datetime(
    c("2019-07-31",
      "2020-07-31", "2021-07-31", "2022-07-31", "2023-07-31", "2024-07-31", "2025-07-31")),
  y_start = -1000,
  y_end = 1000
)


p <- dat %>% 
  mutate(index = if_else(value > 0, "upwelling", "downwelling")) %>% 
  ggplot(aes(timestamp_utc, value, fill = index, col = index)) +
  geom_rect(
    data = summer, inherit.aes = FALSE, fill = "red", alpha = 0.5,
    aes(xmin = x_start, xmax = x_end, ymin = y_start, ymax = y_end)) +
  geom_col() +
  scale_y_continuous("BUI (m^3/s per 100 km)") +
  scale_fill_manual(
    values = c("upwelling" = "#414487FF", "downwelling" = "#7AD151FF")) +
  scale_color_manual(
    values = c("upwelling" = "#414487FF", "downwelling" = "#7AD151FF")) +
  facet_wrap(~station, ncol = 1) +
  theme(
    legend.position = "none", 
    strip.text = element_text(size = 12),
    strip.background = element_rect(fill = "grey50"))
#  panel.spacing.y = unit(5, "lines")) 

p
ggplotly(p, height = 600)

```










