---
title: "Vector Averages"
date: "`r Sys.Date()`"
author: "Danielle Dempsey"
format: 
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
echo: false
editor_options: 
  chunk_output_type: console
---


```{r}
#| include: false

library(dplyr)
library(ggplot2)
#library(plotly)
library(RColorBrewer)

theme_set(theme_light())

v_calculate_components <- function(theta_degrees, v_length) {
  
  v_x <- v_length * sin(theta_degrees * pi / 180)
  v_y <- v_length * cos(theta_degrees * pi / 180)
  
  data.frame(v_x = v_x, v_y = v_y)
}


v_plot_components <- function(
    theta_degrees, v_length = 1, add_components = TRUE
) {
  
  v <- v_calculate_components(theta_degrees, v_length) %>% 
    mutate(
      theta_degrees = ordered(theta_degrees, levels = sort(theta_degrees))
      )
  
  if(nrow(v) < 3) {
    n_theta <- 3
  } else n_theta <- nrow(v)
  
  
  p_v <- ggplot() +
    geom_hline(yintercept = 0, col = "grey60") +
    geom_vline(xintercept = 0, col = "grey60") +
    scale_x_continuous("x", limits = c(-1, 1)) +
    scale_y_continuous("y", limits = c(-1,1)) +
    coord_fixed(ratio = 1)+ 
    geom_segment(
      aes(x = 0, y = 0, xend = v[,1], yend = v[,2], col = v[,3]),
      arrow = arrow(length = unit(0.3, "cm"), type = "closed")
    ) +
    scale_colour_manual("theta (degrees)", values = brewer.pal(n_theta, "Dark2"))
  
  if(isTRUE(add_components)) {
    
    p_v <- p_v +
      # x component
      geom_segment(
        aes(x = 0, xend = v[,1], y = v[,2], yend = v[,2]),
        arrow = arrow(length = unit(0.2, "cm"), type = "closed"),
        linetype = 2,
      ) +
      # y component
      geom_segment(
        aes(x = 0, xend = 0, y = 0, yend = v[,2]),
        arrow = arrow(length = unit(0.2, "cm"), type = "closed"),
        linetype = 2
      )
  }
  
  p_v
}


dat <- data.frame(theta_degrees = c(20, 340), v_length = c(1, 1))

#dat <- cbind(theta_degrees = c(20, 340), v_length = c(1, 1))

v_calculate_average <- function(theta_degrees, v_length) {
  
  v_comp <- v_calculate_components(theta_degrees, v_length) 
  
  v_x_mean <- mean(v_comp$v_x)
  v_y_mean <- mean(v_comp$v_y)
  
  theta <- round(atan2(v_x_mean, v_y_mean) * 180/pi, digits = 2)
  theta <- (360 + theta) %% 360 # maps from (-180 to 180) to (0, 359)
  
  l <- sqrt(v_y_mean^2 + v_x_mean^2)
  
  data.frame(
    v_x = v_x_mean,
    v_y = v_y_mean, 
    theta_degrees = theta, 
    v_length = l
  )
}


# v %>% 
#   plot_ly(x = ~x_start,
#           y = ~y_start) %>% 
#   add_markers() %>% 
#   add_annotations(text = ~angle_deg,
#                   showarrow = TRUE,
#                   xref = "x_start", axref = "x_start",
#                   yref = "y_start", ayref = "y_start",
#                   x = ~x_end,
#                   ax = ~x_start,
#                   y = ~y_end,
#                   ay = ~y_start)

```

# Motivation

Consider two vectors, one pointing to 0 degrees and one pointing to 360 degrees.

These vectors both point North. The average should therefore also point North.

```{r}
#| echo: false

v_plot_components(
  theta_degrees = c(0, 360), v_length = c(1, 0.5), 
  add_components = FALSE
)
```

Taking a simple average:

```{r}
#| echo: true

v_north <- c(0, 360)

mean(v_north)
```

180 degrees points South!

Similarly, consider a vector pointing to 20 degrees and a vector pointing to 340 degrees (-20 degrees).

```{r}
#| echo: false

v_plot_components(
  theta_degrees = c(20, 340), v_length = c(1, 1), 
  add_components = FALSE
)
```

Taking a simple average:

```{r}
#| echo: true

v_north <- c(20, 340)

mean(v_north)
```

180 degrees points South!


We need another way to calculate the average direction.

# Trigonometry Review

## Example 1

Now let's consider a vector of length 1 pointing to the North-East, 45 degrees from vertical.

```{r}
#| echo: false

v_plot_components(
  theta_degrees = 45, v_length = 1, add_components = FALSE
) +
  annotate("text", x = 0.07, y = 0.15, label = "45\u00B0")

```

This can be broken down into 2 components: one along the x-axis, and one along the y-axis.

```{r}

v_plot_components(theta_degrees = 45, v_length = 1) +
  annotate("text", x = 0.07, y = 0.15, label = "45\u00B0")

```

This makes a right-angle triangle, with our original vector as the hypotenuse and the angle theta between the y-axis and our vector.

Now we get to use our high school trigonometry: SOH CAH TOA!

### x component

We use the sine to calculate the length of the x component, since it is opposite from the angle theta (SOH).

$v_x = v_{length} * sin(theta)$

This equation assumes theta is in radians.

Recall: $rad = degrees * pi / 180$

```{r}
#| echo: true
v_x = 1 * sin(45 * pi / 180)

v_x
```

### y component

We use the cosine to calculate the length of the y component, since it is adjacent to the angle theta (CAH).

$v_y = v_{length} * cos(theta)$

```{r}
#| echo: true
v_y = 1 * cos(45 * pi / 180)

v_y
```

Note that `v_x` and `v_y` have the same length. This is because theta = 45 degrees.

If you remember your unit circle, you might have also noticed the component lengths correspond to sqrt(2) / 2.

## Example 2

Now let's consider a vector of length 0.75 pointing 20 degrees from vertical.

```{r}
v_plot_components(theta_degrees = 20, v_length = 0.75) +
  annotate("text", x = 0.07, y = 0.35, label = "20\u00B0")
```


```{r}
#| echo: true
v_x = 0.75 * sin(20 * pi / 180)
v_x

v_y = 0.75 * cos(20 * pi / 180)
v_y
```

## Example 3

This also works for other quadrants. Consider an angle of 170 degrees and length of 1.

```{r}
v_plot_components(theta_degrees = 170, v_length = 1) 
```

```{r}
#| echo: true
v_x = 1 * sin(170 * pi / 180)
v_x

v_y = 1 * cos(170 * pi / 180)
v_y
```


## Example 4

Consider an angle of 230 degrees and length of 0.8.

```{r}
v_plot_components(theta_degrees = 230, v_length = 0.8) 
```

```{r}
#| echo: true
v_x = 0.8 * sin(230 * pi / 180)
v_x

v_y = 0.8 * cos(230 * pi / 180)
v_y
```

## Example 5

Consider an angle of 325 degrees and length of 0.9.

```{r}
v_plot_components(theta_degrees = 325, v_length = 0.9) 
```

```{r}
#| echo: true
v_x = 0.9 * sin(325 * pi / 180)
v_x

v_y = 0.9 * cos(325 * pi / 180)
v_y
```


# Vector Averages

## Example 1

Let's go back to our example with one vector (v1) pointing to 20 degrees and another vector( v2) pointing to 340 degrees (-20 degrees). Let's break them each into their components.

```{r}
#| echo: false

v_plot_components(theta_degrees = c(20, 340), v_length = c(1, 1))
```

The v1 components:
```{r}
#| echo: true
v1_x = 1 * sin(20 * pi / 180)
v1_x

v1_y = 1 * cos(20 * pi / 180)
v1_y
```

The v2 components:
```{r}
#| echo: true
v2_x = 1 * sin(340 * pi / 180)
v2_x

v2_y = 1 * cos(340 * pi / 180)
v2_y
```

Now take the average of the x components and the average of the y components and use these to calculate the new vector.

x component:
```{r}
#| echo: true
v_x <- mean(c(v1_x, v2_x))

v_x
```

y component:
```{r}
#| echo: true
v_y <- mean(c(v1_y, v2_y))

v_y
```

New vector angle:

$theta = atan2(v_x, v_y) * 180 / pi$

New vector length (from Pythagorus):

$l = sqrt(v_x^2 - v_y^2)$

```{r}
theta <- atan2(v_x, v_y) * 180/pi
theta <- (360 + theta) %% 360 # scales to (0, 360)

l <- sqrt(v_y^2 + v_x^2)
```

```{r}
v_plot_components(
  theta_degrees = c(20, 340, theta), v_length = c(1, 1, l),
  add_components = FALSE
)
```


## Example 2

A vector at 0 degrees and a vector at 90 degrees, of equal length. The average vector should be at 45 degrees.

```{r}
v_avg <- v_calculate_average(
  theta_degrees = c(0, 90), v_length = c(1, 1)
)

v_plot_components(
  theta_degrees = c(0, 90, v_avg$theta_degrees), v_length = c(1, 1, v_avg$v_length),
  add_components = FALSE
)
```

What if one vector is longer than the other?

```{r}
v_avg <- v_calculate_average(
  theta_degrees = c(0, 90), v_length = c(1, 0.5)
)

v_plot_components(
  theta_degrees = c(0, 90, v_avg$theta_degrees), 
  v_length = c(1, 0.5, v_avg$v_length),
  add_components = FALSE
)
```

The average vector is "pulled" towards the longer vector.

## Example 3

A vector of length 1 at 340 degrees and a vector of length 0.25 at 190 degrees. The average vector will be weighted by the largest original vector.

```{r}
v_avg <- v_calculate_average(
  theta_degrees = c(340, 190), v_length = c(1, 0.25)
)

v_plot_components(
  theta_degrees = c(340, 190, v_avg$theta_degrees), 
  v_length = c(1, 0.25, v_avg$v_length),
  add_components = FALSE
)
```

## Example 4

This also works for more than 2 vectors.

- v1: 45 degrees, length 1
- v2: 60 degrees, length 1
- v3: 170 degrees, length 1

```{r}

v_avg <- v_calculate_average(
  theta_degrees = c(45, 60, 170), v_length = c(1, 1, 1)
)

v_plot_components(
  theta_degrees = c(45, 60, 170, v_avg$theta_degrees), 
  v_length = c(1, 1, 1, v_avg$v_length),
  add_components = FALSE
)
```

Make v1 and v2 half the length of v3:

```{r}

v_avg <- v_calculate_average(
  theta_degrees = c(45, 60, 170), v_length = c(0.5, 0.5, 1)
)

v_plot_components(
  theta_degrees = c(45, 60, 170, v_avg$theta_degrees), 
  v_length = c(0.5, 0.5, 1, v_avg$v_length),
  add_components = FALSE
)
```

## Example 5

And also vectors from different quadrants

```{r}

v_dat <- data.frame(
  theta_degrees = c(10, 170, 185, 190, 275),
  v_length = c(0.25, 0.5, 0.5, 1, 1)
)

v_avg <- v_calculate_average(v_dat$theta_degrees, v_dat$v_length)

v_plot_components(
  theta_degrees = c(v_dat$theta_degrees, v_avg$theta_degrees), 
  v_length = c(v_dat$v_length, v_avg$v_length),
  add_components = FALSE
)
```





