---
title: "Bakun's Upwelling Index"
date: "`r Sys.Date()`"
author: "Danielle Dempsey"
format: 
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
echo: false
editor_options: 
  chunk_output_type: console
---


```{r}
#| include: false
 
knitr::opts_chunk$set(fig.width = 8, fig.height = 4)

library(dplyr)
library(ggplot2)
library(here)
library(oce)
library(RColorBrewer)

source(here("functions/vector_helpers.R"))

theme_set(theme_light())
```

# Overview

The Bakun Upwelling Index (BUI) is an upwelling index based on Ekman transport, typically expressed in units of $m^3/s$ per 100 m of coastline. Positive values indicate upwelling, and negative values indicate downwelling.

The BUI is calculated from the the component of Ekman transport that is  perpendicular to the coast. If Ekman transport is away from the coast, deep offshore water will flow towards the coast, creating upwelling.

Ekman transport is the net movement of ocean water perpendicular to the wind direction. A combination of wind stress and Coriolis effect cause the water to flow 90° to the right in the Northern Hemisphere (and 90° to the left in the Southern Hemisphere).

Ekman transport is calculated as (e.g., Scrosati and Ellrich, 2023 and references therein):

$M = \tau / (\rho_{water} * f)$, where

$\tau$ is the wind stress:

$\tau = \rho_{air} * C_D * U^2$.

$U$ is the wind speed in m/s, and the remaining parameters are physical constants.

  - $\rho_{air}$: air density (1.28 kg/m$^3$) 
  - $C_D$: the wind drag coefficient (0.0015)
  - $\rho_{water}$: sea water density (1026 kg/m$^3$).
  - $f$: Coriolis parameter, which is dependent on latitude[^1]

[^1]: $f = 2\Omega\sin\phi$, where $\Omega$: Earth's rotation rate (7.292x10$^{-5}s^{-1}$) and $\phi$ is the latitude.

To calculate the BUI, the Ekman flow must be resolved into the along shore and cross-shore components. This can be done in a few different ways. 


# Example

Let's take a simple example of wind blowing at 10 m/s along the coast of Nova Scotia (from 240°, towards 60°). Assume a latitude of 45°.

Since the wind is blowing directly along the shore, we expect the along shore Ekman transport to be zero, and the cross-shore transport to equal the total transport.

```{r}

deg_to_rad <- pi / 180.

wind_speed_m_s <- 10
wind_direction_degree <- 240      # so we know there should be upwelling 
coastline_angle_degree <- 60      # NS coast angle
latitude <- 45

#p_water <- 1026 # density of water
#p_air <- 1.28   # density of air
#cd <- 0.0015    # wind drag coefficient
p_water <- 1025
p_air <- 1.22
cd <- 0.0013
f <- oce::coriolis(latitude) # coriolis parameter
```

```{r}
U_wind <- v_calculate_components(wind_direction_degree, wind_speed_m_s)
U_wind$labels <- "Wind Speed &\nDirection"

v_plot_components(U_wind, lims = 10, pal = "#7570B3")
```

## Method 1

In the first example, we will: 

1. Resolve the u and v (x and y) components of the wind vector.
2. Calculate the corresponding components of $\tau$.
3. Calculate the corresponding components of $M$.
4. Rotate the components of $M$ to be parallel (alongshore) and perpendicular (cross-shore) to the coast.

### 1. Resolve Wind Speed Components

We first calculate the u and v (x and y) components of the wind. Recall:

$u = -(wind\ speed) * sin(theta)$

$v = -(wind\ speed) * cos(theta)$

Where theta is the wind direction in radians, relative to the North.

::: {.callout-note collapse="false"}
### Wind Components

Note we use the negative values of the u and v wind components.

This is because a wind direction of 240° is coming **from** the Southwest, blowing **towards** the Northeast. To make the vector math work, we want the components that point in the direction the wind is blowing **to**.

This is the same as using (wind_direction_degree - 180°).
:::

```{r}
#| echo: true

u = -wind_speed_m_s * sin(wind_direction_degree * deg_to_rad)
v = -wind_speed_m_s * cos(wind_direction_degree * deg_to_rad)

u
v
```

```{r}
v_wind <- v_calculate_components(wind_direction_degree, wind_speed_m_s)

wind_components <- data.frame(
  x = c(u, 0, u),
  y = c(0, v, v),
  labels = c("u", "v", "Wind")
)

v_plot_components(wind_components, lims = 10)
```

### 2. Calculate Wind Stress

Wind stress components are calculated using quadratic drag law: 

$\tau_x = \rho_{air} * C_D * U * u$

$\tau_y = \rho_{air} * C_D * U * v$

$u$ is the component of wind speed in the x direction.

$v$ is the component of wind speed in the y direction.

::: {.callout-tip collapse="false"}
### Pause to think about units:
$kg/m^3 * m^2/s^2 = kg / (m * s^2) = Pa$

Pascals (Pa) is the SI unit for pressure and stress, so this makes sense!
:::

```{r}
#| echo: true

tau_x = p_air * cd * wind_speed_m_s * u
tau_y = p_air * cd * wind_speed_m_s * v

tau_x
tau_y
```

```{r}
tau_components <- data.frame(
  x = c(tau_x, 0), 
  y = c(0, tau_y), 
  labels = c("tau_x", "tau_y")
)

v_plot_components(tau_components, lims = 0.2)
```


### 3. Ekman Transport

Next, we calculate the Ekman transport in in the x and y (u and v) directions. 

Recall:

$M_x = \tau_y / (\rho_{water} * f)$

$M_y = -\tau_x / (\rho_{water} * f)$

::: {.callout-note collapse="false"}
### Ekman Equations

Ekman transport moves water perpendicular to the wind stress - this is why $M_x$ is related to $tau_y$ (and $M_y$ is related to $tau_x$.

In the Northern Hemisphere, the water is moved to the right because of the Coriolis force. This results in the minus sign in $M_y$.
:::


::: {.callout-tip collapse="false"}
### Pause to think about units:
$kg/(m * s^2) / (kg/m^3 * s^-1) = kg/(m * s^2) * (m^3*s)/kg = m^2/s$

*Not sure if this makes sense. Scrosati has units of m3/s. Might need to multiply by a coast length to get volume*
:::

```{r}
#| echo: true

M_x = tau_y / (p_water * f)
M_y = -tau_x / (p_water * f)

M_x
M_y
```

```{r}
M_components <- data.frame(
  x = c(M_x, 0, M_x), 
  y = c(0, M_y, M_y), 
  labels = c("Mx", "My", "M")
)

v_plot_components(M_components, lims = 1.5)
```


### 4. Alongshore and Cross-Shore

We found Ekman transport in the x and y (u and v directions). For the BUI, we are interested in the **cross-shore** Ekman transport (perpendicular to the coast).

We can use the rotation matrix to rotate $M_x$ and $M_y$ so that $M_x$ is parallel to the coast, which means $M_y$ will be perpendicular to the coast.

::: {.callout-note collapse="false"}
### Linear Algebra Review: The rotation matrix

Consider a point $[x, y]$. If the point is rotated $\theta$ degrees, the new coordinates $[x', y']$ are given by:

```{=tex}
\begin{bmatrix} x' \\ y' \end{bmatrix} = \begin{bmatrix} cos(\theta) & -sin(\theta) \\ sin(\theta) & cos(\theta) \end{bmatrix} \begin{bmatrix} x \\ y \end{bmatrix}
```

Positive $\theta$ results in counter-clockwise rotation.

Negative $\theta$ results in clockwise rotation.

Using matrix multiplication:

$x' = cos(\theta)x - sin(\theta)y$

$y' = sin(\theta)x + cos(\theta)y$

Consider a point, p1 at $[1, 0]$ rotated 30° from the horizontal. The new coordinates are:

```{r}
#| echo: true

p1 <- c(1, 0)

x_new <- cos(30 * pi / 180) * 1 - sin(30 * pi / 180) * 0
y_new <- sin(30 * pi / 180) * 1 + cos(30 * pi / 180) * 0

x_new
y_new
```

```{r}

p <- data.frame(
  x = c(1, x_new), 
  y = c(0, y_new), 
  labels = c("p1", "p1'")
)

v_plot_components(p, lims = 1.5) +
 annotate("text", x = 0.35, y = 0.06, label = "30\u00B0") 
```

Now let's do two points, p1 at $[1, 0]$, and p2 at $[0, 1]$ rotated 45° from the horizontal. 

```{r}
#| echo: true

theta <- 45 * pi / 180

# check how matrices are created - fills in columns first
x <-  matrix(data = c("A", "B", "C", "D"), nrow = 2)
x

# matrix of points. First column is p1 and second column is p2
p <- matrix(data = c(1, 0, 0, 1), nrow = 2)

x_new <- cos(theta) * p[1,] - sin(theta) * p[2,]
y_new <- sin(theta) * p[1,] + cos(theta) * p[2,]

x_new
y_new

# using matrix multiplication (make sure to use the %*% operator):
r <- matrix(
  data = c(cos(theta), sin(theta), -sin(theta), cos(theta)), ncol = 2
)

p_r <- r %*% p

p_r
```

```{r}
r_pal <- c("#157A5C", "#D95F02", "#23D19D", "#FDA35D")

p_comp <- data.frame(
  x = c(1, 0,  x_new), 
  y = c(0, 1, y_new), 
  labels = c("p1", "p2", "p1'", "p2'")
)

v_plot_components(p_comp, lims = 1.5, pal = r_pal) 
```


Now let's rotate -45° from the horizontal:

```{r}
#| echo: true

theta <- -45 * pi / 180

r <- matrix(
  data = c(cos(theta), sin(theta), -sin(theta), cos(theta)), ncol = 2
)

p_r <- r %*% p

p_r
```

```{r}

ggplot() +
  geom_hline(yintercept = 0, col = "grey60") +
  geom_vline(xintercept = 0, col = "grey60") +
  scale_x_continuous("x", limits = c(-1.1, 1.1)) +
  scale_y_continuous("y", limits = c(-1.1, 1.1)) +
  coord_fixed(ratio = 1) +
  geom_segment(
    aes(x = 0, y = 0, xend = p[1,], yend = p[2, ], col = "p"),
    arrow = arrow(length = unit(0.3, "cm"), type = "closed")
  ) +
  geom_segment(
    aes(x = 0, y = 0, xend = p_r[1,], yend = p_r[2,], col = "p_rotated"), 
    arrow = arrow(length = unit(0.3, "cm"), type = "closed")
  ) +
  theme(
    legend.position = "inside",
    legend.position.inside = c(0.2, 0.2)
  )
```

This should be the same as rotating 360° - 45° = 315°. 

```{r}
#| echo: true

alpha <- 315 * pi / 180

r <- matrix(
  data = c(cos(alpha), sin(alpha), -sin(alpha), cos(alpha)), ncol = 2
)

p_r <- r %*% p

p_r
```

```{r}

ggplot() +
  geom_hline(yintercept = 0, col = "grey60") +
  geom_vline(xintercept = 0, col = "grey60") +
  scale_x_continuous("x", limits = c(-1.1, 1.1)) +
  scale_y_continuous("y", limits = c(-1.1, 1.1)) +
  coord_fixed(ratio = 1) +
  geom_segment(
    aes(x = 0, y = 0, xend = p[1,], yend = p[2, ], col = "p"),
    arrow = arrow(length = unit(0.3, "cm"), type = "closed")
  ) +
  geom_segment(
    aes(x = 0, y = 0, xend = p_r[1,], yend = p_r[2,], col = "p_rotated"), 
    arrow = arrow(length = unit(0.3, "cm"), type = "closed")
  ) +
  theme(
    legend.position = "inside",
    legend.position.inside = c(0.2, 0.2)
  )
```

:::

### 4. Alongshore and Cross-Shore

Now we can calculate the components that are perpendicular and parallel to the coast by rotating them by the coastline angle $\alpha$. 

:::{.callout-note collapse="false"}
We want to rotate the vectors clockwise, so we actually use -$\alpha$ or 360° - $\alpha$ in the code.

**Is this because the coast is on the left??? Or just by convention?**
:::

$M_{x} = cos(\alpha) * M_x - sin(\alpha) * M_y$

$M_{y} = sin(\alpha) * M_x + cos(\alpha) * M_y$


```{r}

alpha_rad <- (360 - 60) * pi / 180

M_perp1 <- cos(alpha_rad) * M_x + sin(alpha_rad) * M_y
M_para1 <- -sin(alpha_rad) * M_x + cos(alpha_rad) * M_y 

M_perp1
M_para1


r <- matrix(
  data = c(cos(alpha_rad), -sin(alpha_rad), sin(alpha_rad), cos(alpha_rad)), 
  ncol = 2
)

M_mat <- matrix(data = c(M_x, 0, 0, M_y), nrow = 2)

M_mat_r <- r %*% M_mat
M_mat_r
```


```{r}

ggplot() +
  geom_hline(yintercept = 0, col = "grey60") +
  geom_vline(xintercept = 0, col = "grey60") +
  scale_x_continuous("x", limits = c(-0.5, 0.5)) +
  scale_y_continuous("y", limits = c(-0.5, 0.5)) +
  coord_fixed(ratio = 1) +
  geom_segment(
    aes(x = 0, y = 0, xend = M_mat[1,], yend = M_mat[2, ], col = "M"),
    arrow = arrow(length = unit(0.3, "cm"), type = "closed")
  ) +
  geom_segment(
    aes(x = 0, y = 0, xend = M_mat_r[1,], yend = M_mat_r[2,], col = "M_rotated"), 
    arrow = arrow(length = unit(0.3, "cm"), type = "closed")
  ) +
  theme(
    legend.position = "inside",
    legend.position.inside = c(0.8, 0.9)
  )

```


:::{.callout-note collapse="false"}

We have been talking about several different angles. All directions are from the vertical and reported in degrees (they are converted to radians in the code). Recall:

$\theta$ = wind direction  

$\beta$ = Ekman flow direction (before being resolved into x and y components)

$\alpha$ = the angle of the coast **this one might be from horizontal**
:::

# From Total Wind

Overall Ekman transport:

$M = sqrt(M_x^2 + M_y^2)$

Ekman transport direction:

$\beta = atan2(M_x, M_y)$

```{r}
#| echo: true
M = sqrt(M_x^2 + M_y^2)

beta <- round(atan2(M_x, M_y) * 180/pi, digits = 2)
beta <- (360 + beta) %% 360 # maps from (-180 to 180) to (0, 359)

M
beta
```



This is not the same as M for some reason

```{r}

alpha <- 60
theta <- 240

M_check = (1.28 * 0.0015 * 5^2) / (1026 * coriolis(45))

M_check * cos((alpha - theta) * pi / 180)

# new Eck length
sqrt(M_mat_r[1, 1]^2 + M_mat_r[2,1]^2)

```

```{r}
#| message: false

v_plot_components(beta, M, lims = 2) 
```



