---
format: 
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
params:
  round_int: round_int
echo: false
dpi: 600
editor_options: 
  chunk_output_type: console
---

```{r}
#| include: false

library(dplyr)
library(here)
library(mooring)
library(sensorstrings)

source(here("functions/subchunkify.R"))

station_depl <- data.frame(
  station = c("Wedgeport",
              "Big Pond Point",
              "Center Bay",
              "Moose Point 1",
              "Birchy Head",
              "Spry Harbour"),
  depl_date = c("2025-07-31", "2024-08-07", "2024-11-07", 
                "2025-11-20", "2025-12-06", "2024-09-12"),
  mean_current_m_s = c(0.16, 0.295, 0.061, 0.076, 0.075, 0.04),
  max_current_m_s = c(0.471, 0.745, 0.398, 0.493, 0.787, 0.24)
)

```

# Explore Mooring Knockdown

## Assumptions

- Model for one deployment at each station. Other deployment configurations may differ slightly.
- Current speed estimates are from the nearest ADCP. Linear current profile is used for all models.
- In the case where a hobo sensor was attached to the VR2AR sensor, the hobo sensor is not included.
- The bottom float is not included.
- Does not account for biofouling.
- Does not distinguish between VR2AR and VR2AR-X sensors.
- The rope length from the float to the shallowest sensor is hard-coded at 0.05 m.
- If the sounding is shallower than the deepest sensor, the sounding is set to
the deepest depth + 2 m.
  - Moose Point 1 and Center Bay

## Mean Current Speed

```{r}
#| results: asis
#| warning: false
#| fig-height: 5

for (i in 1:nrow(station_depl)) {
  
  m_i <- ss_model_mooring(station_depl[i,]$station, station_depl[i, ]$depl_date)
  
  m_i <- m_i |> 
    segmentize(by = 0.5) |>
    knockdown(u = station_depl[i, ]$mean_current_m_s)
  
  # m[[i]] <- m_i
  
  cat('\n###', station_depl[i,]$station, '\n')
  
  par(mfrow = c(2, 2))
  # plot(m_i, which = "shape", fancy = TRUE, showDepths = TRUE)
  plot(m_i, fancy = TRUE, showDepths = TRUE)
  plot(m_i, which = "knockdown", fancy = TRUE, showDepths = FALSE)
  plot(m_i, which = "tension", fancy = TRUE, showDepths = FALSE)
  plot(m_i, which = "velocity", fancy = TRUE)
  # dev.off()
  
  #plot(m_i, fancy = TRUE, showDepths = TRUE)
   cat('\n')
}

```


## Max Current Speed

```{r}
#| results: asis
#| warning: false
#| fig-height: 5

for (i in 1:nrow(station_depl)) {
  
  m_i <- ss_model_mooring(station_depl[i,]$station, station_depl[i, ]$depl_date)
  
  m_i <- m_i |> 
    segmentize(by = 0.5) |>
    knockdown(u = station_depl[i, ]$max_current_m_s)
  
  cat('\n###', station_depl[i,]$station, '\n')
  
  par(mfrow = c(2, 2))
  plot(m_i, fancy = TRUE, showDepths = TRUE)
  plot(m_i, which = "knockdown", fancy = TRUE, showDepths = FALSE)
  plot(m_i, which = "tension", fancy = TRUE, showDepths = FALSE)
  plot(m_i, which = "velocity", fancy = TRUE)
  
   cat('\n')
}

```

