---
title: "Bakun's Upwelling Index"
date: "`r Sys.Date()`"
author: "Danielle Dempsey"
format: 
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
echo: false
editor_options: 
  chunk_output_type: console
---


```{r}
#| include: false
 
knitr::opts_chunk$set(fig.width = 8, fig.height = 4)

library(dplyr)
library(ggforce)
library(ggplot2)
library(here)
library(oce)
library(RColorBrewer)

source(here("functions/vector_helpers.R"))

theme_set(theme_light())
```

# Overview

The Bakun Upwelling Index (BUI) is an upwelling index based on Ekman transport, typically expressed in units of $m^3/s$ per 100 m of coastline. Positive values indicate upwelling, and negative values indicate downwelling.

The BUI is calculated from the the component of Ekman transport that is  perpendicular to the coast. If Ekman transport is away from the coast, deep offshore water will flow towards the coast, creating upwelling.

Ekman transport is the net movement of ocean water perpendicular to the wind direction. A combination of wind stress and Coriolis effect cause the water to flow 90° to the right in the Northern Hemisphere (and 90° to the left in the Southern Hemisphere).

Ekman transport is calculated as (e.g., Scrosati and Ellrich, 2023 and references therein):

$M = \tau / (\rho_{water} * f)$, where

$\tau$ is the wind stress:

$\tau = \rho_{air} * C_D * U^2$.

$U$ is the wind speed in m/s, and the remaining parameters are physical constants.

  - $\rho_{air}$: air density (1.28 kg/m$^3$) 
  - $\rho_{water}$: sea water density (1026 kg/m$^3$)
  - $C_D$: the wind drag coefficient (0.0015)
  - $f$: Coriolis parameter, which is dependent on latitude[^1]

[^1]: $f = 2\Omega\sin\phi$, where $\Omega$: Earth's rotation rate (7.292x10$^{-5}s^{-1}$) and $\phi$ is the latitude.

To calculate the BUI, we need to calculate the cross-shore component of the Ekman flow. Below are two methods for doing so.


# Example

Let's take a simple example of wind blowing at 10 m/s along the coast of Nova Scotia (from 240°, towards 60°). Assume a latitude of 45°.

Since the wind is blowing directly along the shore, we expect the along shore Ekman transport to be zero, and the cross-shore transport to equal the total transport.

```{r}

deg_to_rad <- pi / 180.

wind_speed_m_s <- 10
wind_direction_degree <- 240      # so we know there should be upwelling 
coastline_angle_degree <- 60      # NS coast angle
latitude <- 45

#p_water <- 1026 # density of water
#p_air <- 1.28   # density of air
#cd <- 0.0015    # wind drag coefficient
p_water <- 1025
p_air <- 1.22
cd <- 0.0013
f <- oce::coriolis(latitude) # coriolis parameter
```

```{r}
U_wind <- v_calculate_components(wind_direction_degree, wind_speed_m_s)
U_wind$labels <- "Wind Speed &\nDirection"

v_plot_components(U_wind, lims = 10, pal = "#7570B3") +
  annotate("text", x = 1, y = 1.5, label = "60\u00B0", col = "#594D21") +
  geom_arc(aes(x0 = 0, y0 = 0, r = 2.75, start = 0, 
               end = coastline_angle_degree * deg_to_rad), col = "#594D21",
  ) +
  annotate("text", x = 0.25, y = -1.5, label = "240\u00B0", col = "#7570B3") +
  geom_arc(aes(x0 = 0, y0 = 0, r = 4, start = 0, 
        end = wind_direction_degree * deg_to_rad), col = "#7570B3",
  ) 
```


## Method 1

In the first example, we will: 

1. Resolve the u and v (x and y) components of the wind vector.
2. Calculate the corresponding components of $\tau$.
3. Calculate the corresponding components of $M$.
4. Rotate the components of $M$ to be parallel (alongshore) and perpendicular (cross-shore) to the coast.

### 1. Resolve Wind Speed Components

We first calculate the u and v (x and y) components of the wind. Recall:

$u = -(wind\ speed) * sin(theta)$

$v = -(wind\ speed) * cos(theta)$

Where theta is the wind direction in radians, relative to the North.

::: {.callout-note collapse="false"}
### Wind Components

Note we use the negative values of the u and v wind components.

This is because a wind direction of 240° is coming **from** the Southwest, blowing **towards** the Northeast. To make the vector math work, we want the components that point in the direction the wind is blowing **to**.

This is the same as using (wind_direction_degree - 180°).
:::

```{r}
#| echo: true

u = -wind_speed_m_s * sin(wind_direction_degree * deg_to_rad)
v = -wind_speed_m_s * cos(wind_direction_degree * deg_to_rad)

u
v
```

```{r}
v_wind <- v_calculate_components(wind_direction_degree, wind_speed_m_s)

wind_components <- data.frame(
  x = c(u, 0, u),
  y = c(0, v, v),
  labels = c("u", "v", "wind")
)

v_plot_components(wind_components, lims = 10)
```

### 2. Calculate Wind Stress

Now we calculate wind stress in the x and y (u and v) directions. Wind stress components are calculated using quadratic drag law: 

$\tau_x = \rho_{air} * C_D * U * u$

$\tau_y = \rho_{air} * C_D * U * v$

where

  - $U$ is the total wind speed
  - $u$ is the component of wind speed in the x direction
  - $v$ is the component of wind speed in the y direction

::: {.callout-tip collapse="false"}
### Pause to think about units:
$kg/m^3 * m^2/s^2 = kg / (m * s^2) = Pa$

Pascals (Pa) is the SI unit for pressure and stress, so this makes sense!
:::

```{r}
#| echo: true

tau_x = p_air * cd * wind_speed_m_s * u
tau_y = p_air * cd * wind_speed_m_s * v

tau_x
tau_y
```

```{r}
tau_components <- data.frame(
  x = c(tau_x, 0, tau_x), 
  y = c(0, tau_y, tau_y), 
  labels = c("tau_x", "tau_y", "tau")
)

v_plot_components(tau_components, lims = 0.2, pal = c("#7570B3", "#1B9E77", "#D95F02"))
```

::: {.callout-tip collapse="false"}
### $\tau$ Check

Check that $sqrt(\tau_x^2 + \tau_y^2) = \rho_{air} * C_D * U^2 = \tau$

```{r}
#| echo: true

# magnitude of tau based on the components
sqrt(tau_x^2 + tau_y^2)

# direct calculation of tau
p_air * cd * wind_speed_m_s^2
```
:::

### 3. Ekman Transport

Next, we calculate the Ekman transport in in the x and y (u and v) directions. 

Recall:

$M_x = \tau_y / (\rho_{water} * f)$

$M_y = -\tau_x / (\rho_{water} * f)$

::: {.callout-note collapse="false"}
### Ekman Equations

Ekman transport moves water perpendicular to the wind stress - this is why $M_x$ is related to $tau_y$ and $M_y$ is related to $tau_x$.

In the Northern Hemisphere, the water is moved to the right because of the Coriolis force. This results in the vector pointing down, so we need the minus sign in $M_y$.
:::


::: {.callout-tip collapse="false"}
### Pause to think about units:
$kg/(m * s^2) / (kg/m^3 * s^-1) = kg/(m * s^2) * (m^3*s)/kg = m^2/s$

Multiply by 100 m to get Bakun's Index, which has units of $m^3/s$ for 100 m of coastline.
:::

```{r}
#| echo: true

M_x = tau_y / (p_water * f)
M_y = -tau_x / (p_water * f)

M_x
M_y
```

```{r}
M_components <- data.frame(
  x = c(M_x, 0, M_x), 
  y = c(0, M_y, M_y), 
  labels = c("Mx", "My", "M")
)

v_plot_components(
  M_components, lims = 1.5,  pal = c("#7570B3", "#1B9E77", "#D95F02")
)
```

### 4. Alongshore and Cross-Shore

We found Ekman transport in the x and y (u and v directions). For the BUI, we are interested in the **cross-shore** Ekman transport (perpendicular to the coast).

We can use the rotation matrix to rotate $M_x$ and $M_y$ to get the $M_perp$ and $M_{para}$ components.

::: {.callout-note collapse="false"}
### Linear Algebra Review: The rotation matrix

Consider a point $[x, y]$. If the point is rotated $\theta$ degrees from horizontal, the new coordinates $[x', y']$ are given by[2]:

```{=tex}
\begin{bmatrix} x' \\ y' \end{bmatrix} = \begin{bmatrix} cos(\theta) & -sin(\theta) \\ sin(\theta) & cos(\theta) \end{bmatrix} \begin{bmatrix} x \\ y \end{bmatrix}
```

Positive $\theta$ results in counter-clockwise rotation.

Negative $\theta$ results in clockwise rotation.

Using matrix multiplication:

$x' = cos(\theta)x - sin(\theta)y$

$y' = sin(\theta)x + cos(\theta)y$

Consider a point, p1 at $[1, 0]$ rotated 30° from the horizontal. The new coordinates are:

```{r}
#| echo: true

p1 <- c(1, 0)

x_new <- cos(30 * pi / 180) * 1 - sin(30 * pi / 180) * 0
y_new <- sin(30 * pi / 180) * 1 + cos(30 * pi / 180) * 0

x_new
y_new
```

```{r}

p <- data.frame(
  x = c(1, x_new), 
  y = c(0, y_new), 
  labels = c("p1", "p1'")
)

v_plot_components(p, lims = 1.5, add_coast = FALSE) +
  annotate("text", x = 0.35, y = 0.08, label = "30\u00B0") +
  geom_arc(aes(x0 = 0, y0 = 0, r = 0.5, start = (90 - 30) * deg_to_rad, 
               end = pi / 2), col = "#7570B3",
  ) 
```

Now let's do two points, p1 at $[1, 0]$, and p2 at $[0, 1]$ rotated 45° from the horizontal. 

```{r}
#| echo: true

theta <- 45 * pi / 180

# check how matrices are created - fills in columns first
x <-  matrix(data = c("A", "B", "C", "D"), nrow = 2)
x

# matrix of points. First column is p1 and second column is p2
p <- matrix(data = c(1, 0, 0, 1), nrow = 2)

x_new <- cos(theta) * p[1,] - sin(theta) * p[2,]
y_new <- sin(theta) * p[1,] + cos(theta) * p[2,]

x_new
y_new

# using matrix multiplication (make sure to use the %*% operator):
r <- matrix(data = c(cos(theta), sin(theta), -sin(theta), cos(theta)), ncol = 2)

r %*% p
```

Note that a positive $\theta$ results in counter-clockwise rotation:

```{r}
r_pal <- c("#157A5C", "#23D19D","#D95F02", "#FDA35D")

p_comp <- data.frame(
  x = c(1, 0, x_new), 
  y = c(0, 1, y_new), 
  labels = c("p1", "p2", "p1'", "p2'")
)

v_plot_components(
  p_comp, lims = 1.5, pal = r_pal, add_coast = FALSE, legend_pos = c(0.15, 0.82)
) +
  annotate("text", x = 0.25, y = 0.08, label = "45\u00B0") +
  geom_arc(aes(x0 = 0, y0 = 0, r = 0.5, start = theta,  end = pi / 2)) +
  annotate("text", x = -0.12, y = 0.3, label = "45\u00B0") +
  geom_arc(aes(x0 = 0, y0 = 0, r = 0.5, start = 0,  end = -theta)) 
```

Now let's rotate -45° from the horizontal:

```{r}
#| echo: true

theta <- -45 * pi / 180

r <- matrix(
  data = c(cos(theta), sin(theta), -sin(theta), cos(theta)), ncol = 2
)

p_r <- r %*% p
```

```{r}
p_comp <- data.frame(
  x = c(1, 0,  p_r[1,]), 
  y = c(0, 1, p_r[2,]), 
  labels = c("p1", "p2", "p1'", "p2'")
)

v_plot_components(
  p_comp, lims = 1.5, pal = r_pal, add_coast = FALSE, legend_pos = c(0.15, 0.82)
) +
  annotate("text", x = 0.25, y = -0.08, label = "45\u00B0") +
  geom_arc(aes(x0 = 0, y0 = 0, r = 0.5, start = pi/2,  end = pi/2 - theta))+
  annotate("text", x = 0.12, y = 0.3, label = "45\u00B0") +
  geom_arc(aes(x0 = 0, y0 = 0, r = 0.5, start = 0, end = -theta)) 
```

:::


Now we can calculate the components that are perpendicular and parallel to the coast by rotating them by the angle $\beta = 360° - \alpha$, where $\alpha$ is the coastline angle. 

:::{.callout-note collapse="false"}
### Rotation Angle

The coastline angle $\alpha$ (and the wind direction, $\theta$) is measured **clockwise** from the **North**.

The rotation matrix measures angles **counter-clockwise** from the **positive x-axis**.

We need to use $\beta = 360° - \alpha$ to rotate the coordinate systems so they match.

```{r}

v_plot_components(NULL, lims = 10) +
  annotate("text", x = 1, y = 1.5, label = "60\u00B0", col = "#594D21") +
  geom_arc(aes(x0 = 0, y0 = 0, r = 2.75, start = 0, 
               end = coastline_angle_degree * deg_to_rad), col = "#594D21",
  ) +
  annotate("text", x = 0, y = -5, label = "360\u00B0 - 60\u00B0", col = "#7570B3") +
  geom_arc(aes(x0 = 0, y0 = 0, r = 4, start = 0, 
        end = -300 * deg_to_rad), col = "#7570B3",
  ) 
```


:::

$M_{perp} = cos(\beta) * M_x - sin(\beta) * M_y$

$M_{para} = sin(\beta) * M_x + cos(\beta) * M_y$


```{r}

beta_rad <- (360 - coastline_angle_degree) * deg_to_rad

M_perp <- cos(beta_rad) * M_x + sin(beta_rad) * M_y
M_para <- -sin(beta_rad) * M_x + cos(beta_rad) * M_y 

M_perp
M_para

r <- matrix(
  data = c(cos(beta_rad), -sin(beta_rad), sin(beta_rad), cos(beta_rad)), 
  ncol = 2
)

M <- matrix(data = c(M_x, M_y), nrow = 2)

M_r <- r %*% M_mat
```


```{r}
M_components <- data.frame(
  x = c(M_perp, 0), 
  y = c(0, M_para), 
  labels = c("M_perp", "M_para")
)

v_plot_components(M_components, lims = 5, add_coast = FALSE) +
  geom_vline(xintercept = 0,   col = "#e2d7b0", linewidth = 1.2)
```

*Note that $M_{para}$ = 0, although the figure still shows the arrow head.


::: {.callout-tip collapse="false"}
### Eckman Check

Check that the magnitude of $M$ is the same when calculated with $M_x$ and $M_y$ as with $M_{perp}$ and $M_para$.

The angles will be different since they are measuring from different coordinate systems.

```{r}
#| echo: false

# from M_x and M_y
sqrt(M_x^2 + M_y^2)

# from M_perp and M_para
sqrt(M_perp^2 + M_para^2)

# from M_x and M_y
gamma <- round(atan2(M_x, M_y) * 180/pi, digits = 2)
(360 + gamma) %% 360 # maps from (-180 to 180) to (0, 359)

# from M_perp and M_para
gamma <- round(atan2(M_para, M_perp) * 180/pi, digits = 2)
(360 + gamma) %% 360 # maps from (-180 to 180) to (0, 359)

```
:::



## Method 2

In the first example, we will: 

1. Resolve the u and v (x and y) components of the wind vector.
2. Calculate the corresponding components of $\tau$.
3. Calculate the corresponding components of $M$.
4. Rotate the components of $M$ to be parallel (alongshore) and perpendicular (cross-shore) to the coast.

## 1. Calculate Wind Stress

Recall $\tau = \rho_{air} * C_D * U^2$.

```{r}
#| echo: true

tau2 = p_air * cd * wind_speed_m_s^2

tau2
```

## 2. Calculate Eckman Transport

Recall $M = \tau / (\rho_{water} * f)$.


```{r}
M2 = tau2 / (p_water * f)
```



## 3. Calculate Cross-Shore Component

$M_{perp} = M * cos(\alpha - \theta)$  

where

  - $\alpha$: coastline angle
  - $\theta$: wind direction angle 

Both angles are measured from the North in radians.

::: {.callout-note collapse="false"}
### Linear Algebra Review: Vector Projection



```{=tex}
\begin{bmatrix} x' \\ y' \end{bmatrix} = \begin{bmatrix} cos(\theta) & -sin(\theta) \\ sin(\theta) & cos(\theta) \end{bmatrix} \begin{bmatrix} x \\ y \end{bmatrix}
```
:::

::: {.callout-note collapse="false"}
### Wind Direction

As above, wind direction is typically recorded as the direction the wind is blowing **from**. We are interested in the direction the wind is blowing **to**. 

Therefore, $\theta$ = (wind_direction_degree - 180°) * pi / 180.
:::





```{r}

U_wind <- v_calculate_components(80, wind_speed_m_s)
U_wind$labels <- "Wind Speed &\nDirection"

v_plot_components(U_wind, lims = 10, pal = "#7570B3")# +
  annotate("text", x = 1, y = 1.5, label = "60\u00B0", col = "#594D21") +
  geom_arc(aes(x0 = 0, y0 = 0, r = 2.75, start = 0, 
               end = coastline_angle_degree * deg_to_rad), col = "#594D21",
  ) +
  annotate("text", x = 0.25, y = -1.5, label = "240\u00B0", col = "#7570B3") +
  geom_arc(aes(x0 = 0, y0 = 0, r = 4, start = 0, 
        end = wind_direction_degree * deg_to_rad), col = "#7570B3",
  ) 
```




