---
title: "Bakun's Upwelling Index"
date: "`r Sys.Date()`"
author: "Danielle Dempsey"
format: 
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
echo: false
editor_options: 
  chunk_output_type: console
---


```{r}
#| include: false
 
knitr::opts_chunk$set(fig.width = 8, fig.height = 4)

library(dplyr)
library(ggplot2)
library(oce)
library(RColorBrewer)

theme_set(theme_light())

v_calculate_components <- function(theta_degrees, v_length) {
  v_x <- v_length * sin(theta_degrees * pi / 180)
  v_y <- v_length * cos(theta_degrees * pi / 180)
  
  data.frame(v_x = v_x, v_y = v_y)
}


v_plot_components <- function(
    theta_degrees, 
    v_length = 1, 
    add_components = TRUE, 
    pal = NULL,
    lims = 1,
    legend_pos = c(0.8, 0.9)
) {
  
  v <- v_calculate_components(theta_degrees, v_length) %>% 
    mutate(
      theta_degrees = ordered(theta_degrees, levels = sort(theta_degrees))
    )
  
  if(nrow(v) < 3) {
    n_theta <- 3
  } else n_theta <- nrow(v)
  
  if(is.null(pal)) pal <- brewer.pal(n_theta, "Dark2")
  
  p_v <- ggplot() +
    geom_hline(yintercept = 0, col = "grey60") +
    geom_vline(xintercept = 0, col = "grey60") +
    scale_x_continuous("x", limits = c(-lims, lims)) +
    scale_y_continuous("y", limits = c(-lims, lims)) +
    coord_fixed(ratio = 1)+ 
    geom_segment(
      aes(x = 0, y = 0, xend = v[,1], yend = v[,2], col = v[,3]),
      arrow = arrow(length = unit(0.3, "cm"), type = "closed")
    ) +
    scale_colour_manual("theta (degrees)", values = pal) +
    theme(
      legend.position = "inside",
      legend.position.inside = c(0.85, 0.9)
    )
  
  if(isTRUE(add_components)) {
    
    p_v <- p_v +
      # x component
      geom_segment(
        aes(x = 0, xend = v[,1], y = v[,2], yend = v[,2]),
        arrow = arrow(length = unit(0.2, "cm"), type = "closed"),
        linetype = 2,
      ) +
      # y component
      geom_segment(
        aes(x = 0, xend = 0, y = 0, yend = v[,2]),
        arrow = arrow(length = unit(0.2, "cm"), type = "closed"),
        linetype = 2
      )
  }
  
  p_v
}
```

# Overview

The Bakun Upwelling Index (BUI) is the component of Ekman transport that is  perpendicular to the coast. Positive values indicate upwelling, and negative values indicate downwelling.

Ekman transport is the net movement of ocean water perpendicular to the wind direction. A combination of wind stress and Coriolis effect cause the water to flow 90° to the right in the Northern Hemisphere (and 90° to the left in the Southern Hemisphere).

Following Scrosati and Ellrich, 2023 (and references therein), Ekman transport is calculated as :

$M = \tau / (\rho_{water} * f)$.

$\tau$ is the wind stress:

$\tau = \rho_{air} * C_D * U^2$,

where:

$\rho_{air}$ is air density (1.28 kg/m$^3$), $C_D$ is the wind drag coefficient (0.0015), and $U$ is the wind speed in m/s.

$\rho_{water}$ is sea water density (1026 kg/m$^3$).

$f$ is the Coriolis parameter, which is dependent on latitude.

# Calculation Example

Let's take a simple example of wind blowing at 5 m/s along the coast of Nova Scotia (from 240°, towards 60°).

```{r}
v_plot_components(
  theta_degrees = c(60, 240), v_length = c(5, 5), 
  add_components = FALSE, lims = 5
) 
```

## 1. Resolve Wind Speed Components

We first calculate the x and y (u and v) components of the wind. Recall:

$ux = (wind\ speed) * sin(theta)$
$vy = (wind\ speed) * cos(theta)$

Where theta is the wind direction in radians, relative to the North.

```{r}
#| echo: true

# wind components relative to x and y --> these were negative in the 
# original code. Probably depends on which vector you use (blowing to or blowing from)
ux = 5 * sin(240 * pi / 180.)
vy = 5 * cos(240 * pi / 180.)

ux
vy
```

```{r}
v_plot_components(
  theta_degrees = 240, v_length = 5, pal = "#D95F02", lims = 5
) +
  theme(
    legend.position = "inside",
    legend.position.inside = c(0.85, 0.9)
  )
```

## 2. Calculate Wind Stress

*I think this was missing in original code*

Recall 

$\tau_x = \rho_{air} * C_D * u_x^2$

$\tau_y = \rho_{air} * C_D * v_y^2$

::: {.callout-tip collapse="false"}
### Pause to think about units:
$kg/m^3 * m^2/s^2 = kg / (m * s^2) = Pa$

Pascals (Pa) is the SI unit for pressure and stress, so this makes sense!
:::

```{r}
#| echo: true

tau_x = 1.28 * 0.0015 * ux^2
tau_y = 1.28 * 0.0015 * vy^2

tau_x
tau_y
```


## 3. Ekman Transport

Next, we calculate the Ekman transport in in the x and y (u and v) directions. Assume a latitude of 45°. 

Recall:

$M_x = \tau_x / (\rho_{water} * f)$

$M_y = \tau_y / (\rho_{water} * f)$

We can use `oce::coriolis()` to calculate $f$ at the latitude of interest.

::: {.callout-tip collapse="false"}
Pause to think about units:
$kg/(m * s^2) / (kg/m^3 * s^-1) = kg/(m * s^2) * (m^3*s)/kg = m^2/s$

*Not sure if this makes sense. Scrosati has units of m3/s. Might need to multiply by a coast length to get volume*
:::

```{r}
#| echo: true

M_x = tau_x / (1026 * coriolis(45))
M_y = tau_y / (1026 * coriolis(45))

M_x
M_y
```

Overall Ekman transport:

$M = sqrt(M_x^2 + M_y^2)$

Ekman transport direction:

$\beta = atan2(M_x, M_y)$

```{r}
#| echo: true
M = sqrt(M_x^2 + M_y^2)

beta <- round(atan2(M_x, M_y) * 180/pi, digits = 2)
beta <- (360 + beta) %% 360 # maps from (-180 to 180) to (0, 359)

M
beta
```

-- this doesn't seem right. Shouldn't it be perpendicular to the wind? So 330 degrees (or 150 degrees)


```{r}
#| message: false

v_plot_components(beta, M, lims = 0.5) +
  theme(
    legend.position = "inside",
    legend.position.inside = c(0.8, 0.9)
  )
```

## 4. Bakun's Index

We found Ekman transport in the x and y (u and v directions), and the overall Ekman transport and direction relative to the vertical.

For the BUI, we are interested in the Ekman transport **perpendicular to the coast**.

We can use the rotation matrix to rotate $M_x$ and $M_y$ so that $M_x$ is parallel to the coast, which means $M_y$ will be perpendicular to the coast.

::: {.callout-note collapse="false"}
### Linear Algebra Review: The rotation matrix

Consider a point $[x, y]$. If the point is rotated $\theta$ degrees, the new coordinates are given by:

```{=tex}
\begin{bmatrix} x' \\ y' \end{bmatrix} = \begin{bmatrix} cos(\theta) & -sin(\theta) \\ sin(\theta) & cos(\theta) \end{bmatrix} \begin{bmatrix} x \\ y \end{bmatrix}
```

Positive $\theta$ results in counter-clockwise rotation.

Negative $\theta$ results in clockwise rotation.

Using matrix multiplication,

$x' = cos(\theta)x - sin(\theta)y$

$y' = sin(\theta)x + cos(\theta)y$

Consider a point, p1 at $[1, 0]$ rotated 30° from the horizontal. The new coordinates are:

```{r}
#| echo: true

p1 <- c(1, 0)

x_new <- cos(30 * pi / 180) * 1 - sin(30 * pi / 180) * 0
y_new <- sin(30 * pi / 180) * 1 + cos(30 * pi / 180) * 0

x_new
y_new
```

```{r}
ggplot() +
  geom_hline(yintercept = 0, col = "grey60") +
  geom_vline(xintercept = 0, col = "grey60") +
  scale_x_continuous("x", limits = c(-1.1, 1.1)) +
  scale_y_continuous("y", limits = c(-1.1, 1.1)) +
  coord_fixed(ratio = 1) +
  geom_segment(
    aes(x = 0, y = 0, xend = x_new, yend = y_new, col = "[0.87, 0.5]"),
    arrow = arrow(length = unit(0.3, "cm"), type = "closed")
  ) +
  geom_segment(
    aes(x = 0, y = 0, xend = 1, yend = 0, col = "[1, 0]"), 
    arrow = arrow(length = unit(0.3, "cm"), type = "closed")
  ) +
  annotate("text", x = 0.25, y = 0.06, label = "30\u00B0") +
  theme(
    legend.title = element_blank(),
    legend.position = "inside",
    legend.position.inside = c(0.8, 0.9)
  )
```

Now let's do two points, p1 at $[1, 0]$, and p2 at $[0, 1]$ rotated 45° from the horizontal. 

```{r}
#| echo: true

alpha <- 45 * pi / 180

# check how matrices are created - filles in columns first
x <-  matrix(data = c("A", "B", "C", "D"), nrow = 2)
x

# matrix of points. First column is p1 and second column is p2
p <- matrix(data = c(1, 0, 0, 1), nrow = 2)

x_new <- cos(45 * pi / 180) * p[1,] - sin(45 * pi / 180) * p[2,]
y_new <- sin(45 * pi / 180) * p[1,] + cos(45 * pi / 180) * p[2,]

x_new
y_new

# using matrix multiplication (make sure to use the %*% operator):
r <- matrix(data = c(cos(alpha), sin(alpha), -sin(alpha), cos(alpha)), ncol = 2)

p_r <- r %*% p

p_r
```

```{r}

ggplot() +
  geom_hline(yintercept = 0, col = "grey60") +
  geom_vline(xintercept = 0, col = "grey60") +
  scale_x_continuous("x", limits = c(-1.1, 1.1)) +
  scale_y_continuous("y", limits = c(-1.1, 1.1)) +
  coord_fixed(ratio = 1) +
  geom_segment(
    aes(x = 0, y = 0, xend = p[1,], yend = p[2, ], col = "p"),
    arrow = arrow(length = unit(0.3, "cm"), type = "closed")
  ) +
  geom_segment(
    aes(x = 0, y = 0, xend = p_r[1,], yend = p_r[2,], col = "p_rotated"), 
    arrow = arrow(length = unit(0.3, "cm"), type = "closed")
  ) +
  theme(
    legend.position = "inside",
    legend.position.inside = c(0.8, 0.1)
  )
```
:::

## 4. Bakun's Index
Now we can calculate the components that are perpendicular and parallel to the coast by rotating them by the coastline angle $\alpha$. 

:::{.callout-note collapse="false"}
We want to rotate the vectors clockwise, so we actually use -$\alpha$ or 360° - $\alpha$ in the code.

**Is this because the coast is on the left??? Or just by convention?**
:::

$M_{x} = cos(\alpha) * M_x - sin(\alpha) * M_y$

$M_{y} = sin(\alpha) * M_x + cos(\alpha) * M_y$


```{r}

alpha <- -60 * pi / 180

r <- matrix(
  data = c(cos(alpha), sin(alpha), -sin(alpha), cos(alpha)), 
  ncol = 2
)

M_mat <- matrix(data = c(M_x, 0, 0, M_y), nrow = 2)

M_mat_r <- r %*% M_mat

```


```{r}

ggplot() +
  geom_hline(yintercept = 0, col = "grey60") +
  geom_vline(xintercept = 0, col = "grey60") +
  scale_x_continuous("x", limits = c(-0.5, 0.5)) +
  scale_y_continuous("y", limits = c(-0.5, 0.5)) +
  coord_fixed(ratio = 1) +
  geom_segment(
    aes(x = 0, y = 0, xend = M_mat[1,], yend = M_mat[2, ], col = "M"),
    arrow = arrow(length = unit(0.3, "cm"), type = "closed")
  ) +
  geom_segment(
    aes(x = 0, y = 0, xend = M_mat_r[1,], yend = M_mat_r[2,], col = "M_rotated"), 
    arrow = arrow(length = unit(0.3, "cm"), type = "closed")
  ) +
  theme(
    legend.position = "inside",
    legend.position.inside = c(0.8, 0.9)
  )

```


:::{.callout-note collapse="false"}

We have been talking about several different angles. All directions are from the vertical and reported in degrees (they are converted to radians in the code). Recall:

$\theta$ = wind direction  

$\beta$ = Ekman flow direction (before being resolved into x and y components)

$\alpha$ = the angle of the coast **this one might be from horizontal**
:::

# From Total Wind

This is not the same as M for some reason

```{r}

alpha <- 60
theta <- 240

M_check = (1.28 * 0.0015 * 5^2) / (1026 * coriolis(45))

M_check * cos((alpha - theta) * pi / 180)

# new Eck length
sqrt(M_mat_r[1, 1]^2 + M_mat_r[2,1]^2)

```



