---
title: "Calculate Bakun's Upwelling Index"
date: "`r Sys.Date()`"
author: "Danielle Dempsey"
format: 
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
echo: true
editor_options: 
  chunk_output_type: console
---


```{r}
#| include: false

knitr::opts_chunk$set(fig.width = 8, fig.height = 4)

library(dplyr)
library(ggplot2)
library(here)
library(RColorBrewer)
library(oce)

source(here("functions/vector_helpers.R"))

theme_set(theme_light())
```


## Example Scenario

The wind is blowing at `r wind_speed_m_s` m/s from 240° along the coast (brown line).

```{r}

deg_to_rad <- pi / 180.

wind_speed_m_s <- 10
wind_direction_degree <- 240 # so we know there should be upwelling (or maybe 240??)
alpha_degree <- 60          # NS coast angle
latitude <- 45

p_water <- 1026 # density of water
p_air <- 1.28   # density of air
cd <- 0.0015    # wind drag coefficient
f <- coriolis(latitude) # coriolis parameter
```


```{r}
#| echo: false

v_plot_components(
  wind_direction_degree, wind_speed_m_s, lims = 10, add_coast = TRUE
) 
```


## Method 1

Calculate Mx and My based on the x and y wind components. 

Rotate using sin and cos equations derived from rotation matrix.

::: {.callout-note collapse="false"}
Note we use the negative values of the x and y wind components.

This is because a wind direction of 225° is coming **from** the Southwest, blowing **towards** the Northeast. To make the vector math work, we want the components that point in the direction the wind is blowing **to**.

This is the same as using (180° + wind_direction_degree).
:::


```{r}
# wind speed components
u_x = -wind_speed_m_s * sin(wind_direction_degree * pi / 180.)
v_y = -wind_speed_m_s * cos(wind_direction_degree * pi / 180.)

v_plot_components(
  180 + wind_direction_degree, wind_speed_m_s, lims = 10, add_coast = TRUE
) 

# wind stress components - might need negatives here
tau_x = p_air * cd * u_x^2
tau_y = p_air * cd * v_y^2

v_plot_components(
  c(0, 90), c(tau_y, tau_x), 
  lims = 0.15, 
  add_coast = TRUE, 
  add_components = FALSE
) 

# Ekman transport components
M_x = tau_x / (p_water * f)
M_y = tau_y / (p_water * f)

v_plot_components(
  c(0, 90), c(M_y, M_x), 
  lims = 1.5, 
  add_coast = TRUE, 
  add_components = FALSE
) 

# Might need to use 360 - alpha here
alpha_rad <- (360-alpha_degree) * deg_to_rad

# rotate Ekman components to perpendicular and parallel to the coast
M_perp1 <- cos(alpha_rad) * M_x + sin(alpha_rad) * M_y
M_para1 <- -sin(alpha_rad) * M_x + cos(alpha_rad) * M_y

v_perp <- v_calculate_components(90 + alpha_degree, M_perp1)
v_para <- v_calculate_components(alpha_degree, M_para1)

v_plot_components(
  c(0, 90), c(M_y, M_x),
  lims = 1.5,
  add_coast = TRUE,
  add_components = FALSE
) +
  geom_segment(
    aes(
      x = 0, y = 0, xend = v_perp[,1], yend = v_perp[,2]),
    arrow = arrow(length = unit(0.3, "cm"), type = "closed")
  ) #+
  geom_segment(
    aes(
      x = 0, y = 0, xend = v_para[,1], yend = v_para[,2]),
    arrow = arrow(length = unit(0.3, "cm"), type = "closed")
  )

M_perp1
M_para1
```

`r knitr::knit_exit()`

## Method 2

Calculate Mx and My based on the x and y wind components. 

Rotate based on rotation matrix.

```{r}

r <- function(alpha_rad) {
  matrix(
    data = c(cos(alpha_rad), -sin(alpha_rad), sin(alpha_rad), cos(alpha_rad)), 
    ncol = 2
  )
}

M_mat <- matrix(data = c(M_x, M_y), ncol = 1)

M_mat_r <- r(alpha_rad) %*% M_mat

M_perp2 = M_mat_r[1,1]
M_para2 = M_mat_r[2,1]
```
## Method 3

Use function from NOAA [oceanview website](https://oceanview.pfeg.noaa.gov/products/upwelling/bakun).

Needs the Ekman components as inputs.

```{r}

upwell <- function(ektrx, ektry, coast_angle) {
  
  pi <- 3.1415927 # don't need this since << pi >> is stored already in R
  degtorad <- pi/180.
  
  alpha <- (360 - coast_angle) * degtorad # this makes the rotation clockwise
  # instead of counter-clockwise
  
  # coefficients for rotation matrix
  s1 <- cos(alpha)
  t1 <- sin(alpha)
  s2 <- -1 * t1
  t2 <- s1
  
  # rotates the eckman transport components to be perpendicular and 
  # parallel to the coast
  perp <- (s1 * ektrx) + (t1 * ektry)
  para <- (s2 * ektrx) + (t2 * ektry)
  
  # not sure why this is divided by 10
  return(perp/10)
}


M_perp3 <- upwell(M_x, M_y, alpha_degree)

M_perp3
```

## Method 4

Find the perpendicular component of M using the cosine between the coast angle and the wind angle (from Scrosati 2023).

```{r}

M = sqrt(M_x^2 + M_y^2)

M_perp4 <- M * cos((alpha_degree - wind_direction_degree) * deg_to_rad)

```

# Compare

```{r}
#| echo: false

data.frame(
  method = c("Method 1", "Method 2", "Method 3", "Method 4"),
  perp_component = c(M_perp1, M_perp2, M_perp3, M_perp4)
)

```




