---
title: "Seasonal Variations and Hurricane Teddy"
author: "Ella Headrick"
date: "`r Sys.Date()`"
format: 
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
echo: false
editor_options: 
  chunk_output_type: console
---

```{r}
#| include: false

library(DT)
library(ggplot2)
library(readr)
library(data.table)     # export csv file
library(dplyr)          # wrangling data
library(here)           # file paths
library(lubridate)      # dates & times
library(plotly)         # interactive figures
library(sensorstrings)  # plot data

library(tidyverse)
library(ggpubr)

library(patchwork)
library(hrbrthemes)
library(scales)

#loading in data sets - CMAR water quality data and ECC climate data (avg at 2 hours and 24 hours)
climate_data_2_hours <- readRDS(("data/climate_data_2_hours.RDS"))
avg_temps_2_hours <- readRDS(("data/2.0_averaging_dataset_2_hours.RDS"))

climate_data_24_hours <- readRDS(("data/climate_data_24_hours.RDS"))
avg_temps_24_hours <- readRDS(("data/2.0_averaging_dataset_24_hours.RDS"))

#joining
joint_data_2_hours <- full_join(climate_data_2_hours, avg_temps_2_hours)
joint_data_24_hours <-full_join(climate_data_24_hours, avg_temps_24_hours)

#saving for later
saveRDS(joint_data_24_hours, file = "data/joint_data_24_hours.rds")
saveRDS(joint_data_2_hours, file = "data/joint_data_2_hours.rds")
```

# Looking at the data together - Hurricane Teddy

```{r}
#reading in data
joint_data_24_hours <- readRDS(("data/data_24_hours.RDS"))
joint_data_2_hours <- readRDS(("data/data_2_hours.RDS"))

```

## Static Figure - 24 hours with dates of interest

```{r}
#| fig-width: 8
#| fig-height: 4

filt.joint_data_24_hours <- joint_data_24_hours %>% filter(ts_round >= as.Date("2020-09-02 00:00:00") & 
           ts_round <= as.Date("2021-01-02 00:00:00"))

ggplot(filt.joint_data_24_hours, aes(x = ts_round, y = avg.air.temp, colour = w_station)) +
  geom_line() 
```

# Static Figure - 2 hours with dates of interest

```{r}
#| fig-width: 8
#| fig-height: 4

filt.joint_data_2_hours <- joint_data_2_hours %>% filter(ts_round >= as.Date("2020-09-02 00:00:00") & 
           ts_round <= as.Date("2021-01-02 00:00:00"))

ggplot(filt.joint_data_2_hours, aes(x = ts_round, y = avg.air.temp, colour = w_station)) +
  geom_line()

```

# Interactive Figure - weather station data - 24 hours

```{r}
p24 <- ggplot(filt.joint_data_24_hours, aes(x = ts_round, y = avg.air.temp, colour = w_station)) +
  geom_line()

ggplotly(p24)
```

# Interactive figure - weather station data - 2 hours

```{r}
p2 <- ggplot(filt.joint_data_2_hours, aes(x = ts_round, y = avg.air.temp, colour = w_station)) +
  geom_line()

ggplotly(p2)
```

```{r}
#re-reading in different data to transform and plot (in case its gone from the environment)
avg_temps_2_hours <- readRDS(("data/2.0_averaging_dataset_2_hours.RDS"))
avg_temps_24_hours <- readRDS(("data/2.0_averaging_dataset_24_hours.RDS"))

#transforming the data to make it a little easier to work with 
tr.avg_temp_2 <- avg_temps_2_hours %>% 
  pivot_longer(cols = c(3:5), names_to = "sensor_depth", values_to = "value") %>%
  filter(ts_round >= as.Date("2020-09-02 00:00:00") & 
           ts_round <= as.Date("2021-01-02 00:00:00"))

tr.avg_temp_24 <- avg_temps_24_hours %>% 
  pivot_longer(cols = c(3:5), names_to = "sensor_depth", values_to = "value") %>%
  filter(ts_round >= as.Date("2020-09-02 00:00:00") & 
           ts_round <= as.Date("2021-01-02 00:00:00"))

```


# Interactive figure - site data - 24 hours - Moose Point 1

```{r}
d24 <- tr.avg_temp_24 %>% 
  filter(station == "Moose Point 1") %>% 
  ggplot(aes(x = ts_round, y = value, colour = sensor_depth)) +
  geom_line() +
  labs( 
    y = "temp (degrees C)")

ggplotly(d24)
```

# Interactive figure - site data - 2 hours - Moose Point 1

```{r}
d2 <- tr.avg_temp_2 %>% 
  filter(station == "Moose Point 1") %>% 
  ggplot(aes(x = ts_round, y = value, colour = sensor_depth)) +
  geom_line() +
  labs( 
    y = "temp (degrees C)")

ggplotly(d2)
```


# plotting wind speed and temperature against eachother

```{r}
climate_data_24 <- readRDS(("data/climate_data_24_hours.RDS"))

climate_data_24 %>% 
  filter(w_station == "BEAVER ISLAND (AUT)") %>%
  filter(ts_round >= as.Date("2020-09-02 00:00:00") & 
           ts_round <= as.Date("2021-01-02 00:00:00")) %>% 
  ggplot(aes(x=ts_round)) +
  geom_line(aes(y=avg.air.temp, colour = "avg.air.temp")) + 
  geom_line(aes(y=avg.wind.speed_kmph, colour = "avg.wind.speed_kmph")) +
  scale_y_continuous(name = "avg temperature(degrees C))", 
                     sec.axis = sec_axis(~., name = "avg wind speed (kmph)")) +
  scale_colour_manual(values=c("avg.air.temp"="red", "avg.wind.speed_kmph"="blue"))


W_T <- climate_data_24 %>% 
  filter(w_station == "BEAVER ISLAND (AUT)") %>%
  filter(ts_round >= as.Date("2020-09-02 00:00:00") & 
           ts_round <= as.Date("2021-01-02 00:00:00")) %>% 
  ggplot(aes(x=ts_round)) +
  geom_line(aes(y=avg.air.temp, colour = "avg.air.temp")) + 
  geom_line(aes(y=avg.wind.speed_kmph, colour = "avg.wind.speed_kmph")) +
  scale_y_continuous(name = "avg temperature(degrees C))", 
                     sec.axis = sec_axis(~., name = "avg wind speed (kmph)")) +
  scale_colour_manual(values=c("avg.air.temp"="red", "avg.wind.speed_kmph"="blue"))


ggplotly(W_T)
```


# correlation between wind speed and water temperatature at a site (2m - Spry Harbour)

```{r}

filt.2 <- filt.joint_data_2_hours %>% 
  filter(w_station == "BEAVER ISLAND (AUT)" , station == "Spry Harbour")

ggscatter(filt.2, x = "avg.wind.speed_kmph", y = "depth_2", 
          add = "reg.line", conf.int = TRUE, 
          cor.coef = TRUE, cor.method = "pearson",
          xlab = "wind speed (kmph)", ylab = "air temp (degrees C)")


c <- ggscatter(filt.2, x = "avg.wind.speed_kmph", y = "depth_2", 
          add = "reg.line", conf.int = TRUE, 
          cor.coef = TRUE, cor.method = "pearson",
          xlab = "wind speed (kmph)", ylab = "depth_2 (temp in degrees C)")

ggplotly(c)
```


# Seasonal Variability

```{r}
#reading in data
joint_data_24_hours <- readRDS(("data/data_24_hours.RDS"))
joint_data_2_hours <- readRDS(("data/data_2_hours.RDS"))

```

## one year of data

```{r}

joint_data_24_hours %>% 
filter(ts_round >= as.Date("2020-01-01 00:00:00") & 
           ts_round <= as.Date("2020-12-31 00:00:00")) %>% 
  ggplot(aes(x = ts_round, y = avg.air.temp, colour = w_station)) +
  geom_line()
  
```

##winter to spring (jan-june)

```{r}

joint_data_24_hours %>% 
filter(ts_round >= as.Date("2020-01-01 00:00:00") & 
           ts_round <= as.Date("2020-06-30 00:00:00")) %>% 
  ggplot(aes(x = ts_round, y = avg.air.temp, colour = w_station)) +
  geom_line()
  
```


##summer to fall (july-december)

```{r}

joint_data_24_hours %>% 
filter(ts_round >= as.Date("2020-06-01 00:00:00") & 
           ts_round <= as.Date("2020-12-31 00:00:00")) %>% 
  ggplot(aes(x = ts_round, y = avg.air.temp, colour = w_station)) +
  geom_line()


```


#air temperature and water temperature - spry harbour

```{r}

joint_data_24_hours %>% 
  filter(w_station == "BEAVER ISLAND (AUT)", station == "Spry Harbour") %>%
  filter(ts_round >= as.Date("2020-01-01 00:00:00") & 
           ts_round <= as.Date("2021-12-31 00:00:00")) %>% 
  ggplot(aes(x=ts_round)) +
  geom_line(aes(y=avg.air.temp, colour = "avg.air.temp")) + 
  geom_line(aes(y=depth_2, colour = "depth_2")) +
  geom_line(aes(y=depth_5, colour = "depth_5")) +
  geom_line(aes(y=depth_10, colour = "depth_10")) +
  scale_y_continuous(name = "avg temperature(degrees C))", 
                     sec.axis = sec_axis(~., name = "temp at 2m (degrees C)")) +
  scale_colour_manual(values=c("avg.air.temp"="red", 
                               "depth_2"="lightblue", 
                               "depth_5"="blue",
                               "depth_10"= "darkblue"))


x <- joint_data_24_hours %>% 
  filter(w_station == "BEAVER ISLAND (AUT)", station == "Spry Harbour") %>%
  filter(ts_round >= as.Date("2020-01-01 00:00:00") & 
           ts_round <= as.Date("2021-12-31 00:00:00")) %>% 
  ggplot(aes(x=ts_round)) +
  geom_line(aes(y=avg.air.temp, colour = "avg.air.temp")) + 
  geom_line(aes(y=depth_2, colour = "depth_2")) +
  geom_line(aes(y=depth_5, colour = "depth_5")) +
  geom_line(aes(y=depth_10, colour = "depth_10")) +
  scale_y_continuous(name = "avg temperature(degrees C))", 
                     sec.axis = sec_axis(~., name = "temp at 2m (degrees C)")) +
  scale_colour_manual(values=c("avg.air.temp"="red", 
                               "depth_2"="lightblue", 
                               "depth_5"="blue",
                               "depth_10"= "darkblue"))

ggplotly(x)

```





# correlation between wind air tempertaure and water temperature at depth "X"

## Spry Harbour station - beaver island weather station


# at 2 meters


#at 5 meters


#at 10 meters



## Moose Point 1 station - hart island weather station


