---
title: "Calculate Bakun's Upwelling Index"
date: "`r Sys.Date()`"
author: "Danielle Dempsey"
format: 
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
echo: true
editor_options: 
  chunk_output_type: console
---


```{r}
#| include: false

library(ggplot2)
library(oce)
```


## Define Common Parameters
```{r}
deg_to_rad <- pi / 180.

wind_speed_m_s <- 5
wind_direction_degree <- 60 # so we know there should be upwelling (or maybe 240??)
alpha_degree <- 60          # NS coast angle

p_water <- 1026 # density of water
p_air <- 1.28   # density of air
cd <- 0.0015    # wind drag coefficient
```


## Method 1

Calculate Mx and My based on the x and y wind components. 

Rotate using sin and cos equations derived from rotation matrix.

```{r}
# wind speed components
u_x = wind_speed_m_s * sin(wind_direction_degree * pi / 180.)
v_y = wind_speed_m_s * cos(wind_direction_degree * pi / 180.)

# wind stress components
tau_x = p_air * cd * u_x^2
tau_y = p_air * cd * v_y^2

# Ekman transport components
M_x = tau_x / (p_water * coriolis(45))
M_y = tau_y / (p_water * coriolis(45))

# Might need to use 360 - alpha here
alpha_rad <- alpha_degree * deg_to_rad

# rotate Ekman components to perpendicular and parallel to the coast
M_perp1 <- cos(alpha_rad) * M_x + sin(alpha_rad) * M_y
M_para1 <- -sin(alpha_rad) * M_x + cos(alpha_rad) * M_y

M_perp1
M_para1
```
```{r}

ggplot() +
  geom_hline(yintercept = 0, col = "grey60") +
  geom_vline(xintercept = 0, col = "grey60") +
  scale_x_continuous("x", limits = c(-0.5, 0.5)) +
  scale_y_continuous("y", limits = c(-0.5, 0.5)) +
  coord_fixed(ratio = 1) +
  geom_segment(
    aes(x = 0, y = 0, xend = M_x, yend = M_y, col = "M"),
    arrow = arrow(length = unit(0.3, "cm"), type = "closed")
  ) +
  geom_segment(
    aes(x = 0, y = 0, xend = M_perp1, yend = M_para1, col = "M_rotated"), 
    arrow = arrow(length = unit(0.3, "cm"), type = "closed")
  ) 
```

## Method 2

Calculate Mx and My based on the x and y wind components. Rotate based on rotation matrix.

```{r}

r <- function(alpha_degree) {
  alpha <- alpha_degree * pi / 180
  matrix(data = c(cos(alpha), sin(alpha), -sin(alpha), cos(alpha)), ncol = 2)
}

M_mat <- matrix(data = c(M_x, 0, 0, M_y), nrow = 2)

M_mat_r <- r(alpha_degree) %*% M_mat

M_perp2.1 = sqrt(M_mat_r[1,1]^2 + M_mat_r[2,1]^2)
M_perp2.2 = sqrt(M_mat_r[1,2]^2 + M_mat_r[2,2]^2)

```
## Method 3

Use function from NOAA [oceanview website](https://oceanview.pfeg.noaa.gov/products/upwelling/bakun).

Needs the Ekman components as inputs.

```{r}

upwell <- function(ektrx, ektry, coast_angle) {
  
  pi <- 3.1415927 # don't need this since << pi >> is stored already in R
  degtorad <- pi/180.
  
  alpha <- (360 - coast_angle) * degtorad # this makes the rotation clockwise
  # instead of counter-clockwise
  
  # coefficients for rotation matrix
  s1 <- cos(alpha)
  t1 <- sin(alpha)
  s2 <- -1 * t1
  t2 <- s1
  
  # rotates the eckman transport components to be perpendicular and 
  # parallel to the coast
  perp <- (s1 * ektrx) + (t1 * ektry)
  para <- (s2 * ektrx) + (t2 * ektry)
  
  # not sure why this is divided by 10
  return(perp/10)
}


M_perp3 <- upwell(M_x, M_y, 300)

M_perp3
```

## Method 4

Find the perpendicular component of M using the cosine between the coast angle and the wind angle (from Scrosati 2023).

```{r}

M = sqrt(M_x^2 + M_y^2)

M_perp4 <- M * cos((alpha_degree - wind_direction_degree) * deg_to_rad)

```

# Compare

```{r}
#| echo: false

# c("Method 1" = M_perp1, 
#   "Method 2.1" = M_perp2.1,
#   "Method 2.2" = M_perp2.2,
#   "Method 3" = M_perp3,
#   "Method 4" = M_perp4
#   )

data.frame(
  method = c("Method 1", "Method 2.1", "Method 2.2", "Method 3", "Method 4"),
  perp_component = c(M_perp1, M_perp2.1, M_perp2.2, M_perp3, M_perp4)
)

```




