---
title: "Interpolation"
date: "`r Sys.Date()`"
author: "Danielle Dempsey"
format: 
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
echo: true
fig-height: 4
editor_options: 
  chunk_output_type: console
---


```{r}
#| include: false

library(dplyr)
library(ggplot2)
library(imputeTS)
library(lubridate)
library(purrr)
library(tidyr)

theme_set(theme_light())

A <- 10
per <- 365
phi <- 0
C <- 15
```


Review how `imputeTS::na_seadec()` handles a full time series of missing values.

# Summary 

  - `imputeTS::na_seadec()` will give a `Warning` when there are fewer than 3 non-NA data points in the time series.
  - BUT `imputeTS::na_seadec()` DOES NOT RESPECT `dplyr::group_by()` groups. It will silently use data from one group to fill in the other group.
  - To properly interpolate for multiple groups at a time, break the dataframe into lists (instead of groups) and use `purrr::map()` to apply `na_seadec()` to each list.
  
  

# One Group

## All `NA`

Results in a `Warning`.

```{r}

dat <- data.frame(
  timestamp = seq(as_date("2025-01-01"), as_date("2026-12-31")),
  value = NA
) 

dat_imp <- ts(dat)

imp <- na_seadec(dat_imp, algorithm = "interpolation", find_frequency = TRUE)

```

## Data Values

```{r}
dat1 <- data.frame(
  timestamp = seq(as_date("2025-01-01"), as_date("2026-12-31"))
) |> 
  mutate(value = A * sin((2 * pi * as.numeric(timestamp)) / per  + phi)) 

dat1[20:45, "value"] <- NA
dat1[471:571, "value"] <- NA

ggplot(dat1, aes(timestamp, value)) +
  geom_line() 

dat_imp1 <- dat1 |> 
  ts() |> 
  na_seadec(algorithm = "interpolation", find_frequency = TRUE)

dat1$value_imp <- dat_imp1[, 2]

dat1 <- dat1 |> 
  mutate(imp = if_else(is.na(value), TRUE, FALSE))

ggplot(dat1, aes(timestamp, value_imp, col = imp)) +
  geom_point() 

```

## Bonus Check: Values not in Chronological Order

  - `ggplot()` figure is correct because `ggplot2` recognizes `timestamp` is a `Date` object and re-orders the values accordingly.
  - HOWEVER, `ts()` converts the data frame into a timeseries object based on the row order.
  - So if the rows are not in chronological order, there may not be a seasonal trend to decompose.
  - The interpolation will not be correct.

```{r}
dat1_1 <- data.frame(
  timestamp = seq(as_date("2025-01-01"), as_date("2026-12-31"))
) |> 
  mutate(value = A * sin((2 * pi * as.numeric(timestamp)) / per  + phi)) 

dat1_1[20:45, "value"] <- NA
dat1_1[471:571, "value"] <- NA

dat1_1 <- dat1[sample(1:nrow(dat1_1)), ]

# looks the same as the above because ggplot knows timestamp is a Date object
ggplot(dat1_1, aes(timestamp, value)) +
  geom_line() 

dat_imp1_1 <- dat1_1 |> 
  ts() |> 
  na_seadec(algorithm = "interpolation", find_frequency = TRUE)

dat1_1$value_imp <- dat_imp1_1[, 2]

dat1_1 <- dat1_1 |> 
  mutate(imp = if_else(is.na(value), TRUE, FALSE))

ggplot(dat1_1, aes(timestamp, value_imp, col = imp)) +
  geom_point() 
```

# Two Groups

## All `NA` for 1 Group 

### Using `dplyr::group_by()`

This silently gives the wrong answer!

`imputeTS::na_secdec()` does not respect `dplyr::group_by()`.

```{r}

dat2 <- data.frame(
  timestamp = seq(as_date("2025-01-01"), as_date("2026-12-31")),
  variable_1 = NA
) |> 
  mutate(variable_2 = A * sin((2 * pi * as.numeric(timestamp)) / per  + phi)) |> 
  pivot_longer(cols = contains("variable"), values_to = "value", names_to = "variable") |> 
  arrange(timestamp)
 
dat2[75:125, "value"] <- NA
dat2[1000:1200, "value"] <- NA

ggplot(dat2, aes(timestamp, value)) +
  geom_line() +
  facet_wrap(~variable, ncol = 1)

dat_imp2 <- dat2 |> 
  ts() |> 
  # doesn't give an error or warning!
  # na_secdec does not respect group_by
  na_seadec(algorithm = "interpolation", find_frequency = TRUE) 

dat2$value_imp <- dat_imp2[, 3]

dat2 <- dat2 |> 
  mutate(imp = if_else(is.na(value), TRUE, FALSE)) 

ggplot(dat2, aes(timestamp, value_imp, col = imp)) +
  geom_point() +
  facet_wrap(~variable, ncol = 1)
```

### Using `purrr::map()`

Gives a `Warning` message for groups with too many NAs.

```{r}
dat3 <- data.frame(
  timestamp = seq(as_date("2025-01-01"), as_date("2026-12-31")),
  variable_1 = NA
) |> 
  mutate(variable_2 = A * sin((2 * pi * as.numeric(timestamp)) / per  + phi)) |> 
  pivot_longer(cols = contains("variable"), values_to = "value", names_to = "variable") |> 
  arrange(variable)


dat3[750:775, "value"] <- NA
dat3[1200:1300, "value"] <- NA

ggplot(dat3, aes(timestamp, value)) +
  geom_line() +
  facet_wrap(~variable, ncol = 1)

# gives a warning message!
dat3_imp <- dat3 |> 
  group_split(variable) |> # splits into lists based on a variable
  map(
    .f = \(x) ts(x) |> 
      na_seadec(algorithm = "interpolation", find_frequency = TRUE) |> 
      data.frame()
  ) |> 
  list_rbind() 

dat3$value_imp <- dat3_imp$value 

dat3 <- dat3 |> 
  mutate(imp = if_else(is.na(value), TRUE, FALSE))

ggplot(dat3, aes(timestamp, value_imp, col = imp)) +
  geom_point() +
  facet_wrap(~variable, ncol = 1)

```


## Data Values: `purrr::map()`

```{r}

dat4 <-  data.frame(
  timestamp = seq(as_date("2025-01-01"), as_date("2026-12-31"))) |> 
  mutate(
    variable_1 = A * sin((2 * pi * as.numeric(timestamp)) / per  + phi),
    variable_2 = 5 * sin((2 * pi * as.numeric(timestamp)) / per  + 25)
  ) |> 
  pivot_longer(cols = contains("variable"), values_to = "value", names_to = "variable") |> 
  arrange(variable, timestamp)

dat4[100:175, "value"] <- NA
dat4[300:335, "value"] <- NA

dat4[750:775, "value"] <- NA
dat4[1200:1300, "value"] <- NA

ggplot(dat4, aes(timestamp, value)) +
  geom_line() +
  facet_wrap(~variable, ncol = 1)

dat4_imp <- dat4 |> 
  group_split(variable) |> # splits into lists based on a variable
  purrr::map(
    .f = \(x) ts(x) |> 
      na_seadec(algorithm = "interpolation", find_frequency = TRUE) |> 
      data.frame()
  ) |> 
  list_rbind() 

dat4$value_imp <- dat4_imp$value 

dat4 <- dat4 |> 
  mutate(imp = if_else(is.na(value), TRUE, FALSE))

ggplot(dat4, aes(timestamp, value_imp, col = imp)) +
  geom_point() +
  facet_wrap(~variable, ncol = 1)

```

# Does `padr::pad()` change anything?


```{r}

dat5 <- dat4 |> 
  select(-c(value_imp, imp)) |> 
  filter(!is.na(value))

sum(is.na(dat5$value))
nrow(dat5)

ggplot(dat5, aes(timestamp, value)) +
  geom_point(size = 0.5) +
  facet_wrap(~variable, ncol = 1)

# add NAs back in with padr
dat5 <- dat5 |> 
  group_by(variable) |> 
  padr::pad() |> 
  ungroup()

sum(is.na(dat5$value))
nrow(dat5)

dat5_imp <- dat5 |> 
  group_split(variable) |> # splits into lists based on a variable
  purrr::map(
    .f = \(x) ts(x) |> 
      na_seadec(algorithm = "interpolation", find_frequency = TRUE) |> 
      data.frame()
  ) |> 
  list_rbind() 

dat5$value_imp <- dat5_imp$value 

dat5 <- dat5 |> 
  mutate(imp = if_else(is.na(value), TRUE, FALSE))

ggplot(dat5, aes(timestamp, value_imp, col = imp)) +
  geom_point() +
  facet_wrap(~variable, ncol = 1)

```










